Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-12-17 analysis/paper/paper.qmd
updates

df_long <- df %>%
  pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
  group_by(category) %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(
    category = factor(category, 
                      levels = c("spheroids","Polys_and_Multi_cores", "cores", "total")),
    material = factor(material, 
                      levels = c("Lava", "Dolomite", "Chert", "Other"))
  )

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Cores",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )) %>% 
         filter(category != "Total"), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#b2abd2",
    "Cores" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.88, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black"),
    legend.box.margin = margin(0, 10, 0, 10)
  )

df_long <- df_long %>%
  mutate(text = paste0("n = ", count, ", ", round(prop, 1), "%"))

lava_total_text <- df_long %>% filter(material == "Lava", category == "total") %>% pull(text)
dolomite_total_text <- df_long %>% filter(material == "Dolomite", category == "total") %>% pull(text)
chert_total_text <- df_long %>% filter(material == "Chert", category == "total") %>% pull(text)
other_total_text <- df_long %>% filter(material == "Other", category == "total") %>% pull(text)

lava_spheroids_text <- df_long %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
dolomite_spheroids_text <- df_long %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
chert_spheroids_text <- df_long %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)

lava_multi_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
dolomite_multi_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
chert_multi_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
other_multi_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)

@@

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 248, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). Among these, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 pieces, accounting for 1.3% of the total lithic assemblage. They are all faceted spheroids with intensive flake scars on the surface. Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.

@@

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Cores",
                      "spheroid" = "Spheroid",
                      "multi_poly" = "Polys and Multi cores"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Cores" = "#b2abd2", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  guides(fill = "none")

@@

ggsave(
  filename = here("analysis/paper/figures/basic_characteristics.png"),
  plot = p_combined,
  width = 9,   
  height = 4,  
  dpi = 300
)

knitr::include_graphics(here("analysis/paper/figures/basic_characteristics.png"))
```

@@

```{r, echo=FALSE}
#| label: fig-benn-ternary-plot
#| fig-cap: "Ternary plot showing flake scar orientation analysis"

benn.output$Z <- 1 - benn.output$EL - benn.output$IS

@@

p_benn_ternary <- 
ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
  geom_mask() + 
  geom_point(size = 3) + 
  theme_bw() +
  labs(
    x = "Planar", 
    y = "Isotropic", 
    yarrow = "Isotropy Ratio", 
    z = "Linear", 
    zarrow = "Elongation Ratio"
  ) +
  theme_showarrows() + 
  theme_clockwise() + 
  #geom_text(position = pn, aes(label = type), check_overlap = FALSE, size = 3) +
  scale_color_manual(values = c(
    "multifacial" = "#b2abd2",
    "polyhedron" = "#5e3c99",
    "subspheroid" = "#fdb863",
    "spheroid" = "#e66101"
  ))

ggsave(here("analysis/paper/figures/benn_ternary_plot.jpg"), 
       plot = p_benn_ternary,
       width = 10, 
       height = 6, 
       dpi = 300)

knitr::include_graphics(here("analysis/paper/figures/benn_ternary_plot.jpg"))

>>>>

# 转换数据格式 - 按核心类型分组计算原料比例
df_long <- df %>%
  # 移除total列，因为我们要计算的是每种核心类型内的原料比例
  select(-total) %>%
  pivot_longer(cols = -material, names_to = "core_type", values_to = "count") %>%
  # 按核心类型分组计算百分比
  group_by(core_type) %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(
    core_type = factor(core_type, 
                      levels = c("spheroids", "Polys_and_Multi_cores", "cores"),
                      labels = c("Spheroids", "Polys and Multi cores", "Cores")),
    material = factor(material, 
                      levels = c("Lava", "Dolomite", "Chert", "Other"))
  )

# 正确计算每种核心类型的总数
total_by_core_type_correct <- df_long %>%
  group_by(core_type) %>%
  summarise(total_count = sum(count)) %>%
  ungroup()

# 查看正确总数
print(total_by_core_type_correct)

# 创建带数量的横坐标标签（使用正确总数）
core_type_labels_correct <- c(
  "Spheroids" = paste0("Sub/ Spheroids \n(n = ", 
                       total_by_core_type_correct$total_count[1], ")"),
  "Polys and Multi cores" = paste0("Polyhedrons/ Multifacils\n(n = ", 
                                   total_by_core_type_correct$total_count[2], ")"),
  "Cores" = paste0("Cores\n(n = ", 
                   total_by_core_type_correct$total_count[3], ")")
)

# 绘图
# 绘图
p_raw_material <- ggplot(df_long, 
       aes(x = core_type, 
           y = prop, 
           fill = material)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +

  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Raw Material"  # 添加图例标题
  ) +

  # 添加带正确数量的横坐标标签
  scale_x_discrete(labels = core_type_labels_correct) +

  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.05))
  ) +

  scale_fill_manual(
    values = c(
      "Lava" = "orange",
      "Dolomite" = "#b2abd2",
      "Chert" = "gray",
      "Other" = "gray40"
    )
  ) +

  theme_minimal(base_size = 10) +
  theme(
    # 图例放在右上角
    legend.position = c(0.95, 0.95),
    legend.justification = c(1, 1),
    legend.background = element_rect(fill = "white", color = "black", size = 0.3),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.1, "cm"),
    legend.margin = margin(3, 3, 3, 3),
    legend.title = element_text(size = 9),  # 调整图例标题大小

    # 其他样式
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9, lineheight = 0.8)
  )

# 显示图形
print(p_raw_material)

df_long1 <- df %>%
  pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
  group_by(category) %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(
    category = factor(category, 
                      levels = c("spheroids","Polys_and_Multi_cores", "cores", "total")),
    material = factor(material, 
                      levels = c("Lava", "Dolomite", "Chert", "Other"))
  )

df_long1 <- df_long1 %>%
  mutate(text = paste0("n = ", count, ", ", round(prop, 1), "%"))

lava_total_text <- df_long1 %>% filter(material == "Lava", category == "total") %>% pull(text)
dolomite_total_text <- df_long1 %>% filter(material == "Dolomite", category == "total") %>% pull(text)
chert_total_text <- df_long1 %>% filter(material == "Chert", category == "total") %>% pull(text)
other_total_text <- df_long1 %>% filter(material == "Other", category == "total") %>% pull(text)

lava_spheroids_text <- df_long1 %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
dolomite_spheroids_text <- df_long1 %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
chert_spheroids_text <- df_long1 %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
other_spheroids_text <- df_long1 %>% filter(material == "Other", category == "spheroids") %>% pull(text)

lava_multi_text <- df_long1 %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
dolomite_multi_text <- df_long1 %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
chert_multi_text <- df_long1 %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
other_multi_text <- df_long1 %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)

@@

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017] , the assemblage was classified into four general categories: flaked pieces, detached pieces, pounded pieces, and unmodified materials. The numerical distribution and proportional composition of these categories are provided in the supplementary material. Within the category of flaked pieces, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 artifacts, accounting for 1.3% of the total lithic assemblage. They are all faceted stone tools with intensive flake scars on the surface. Within this continuum, spheroids show a more symmetrical morphology and more extensive flake scar coverage, while subspheroids are characterized by lower symmetry and a less intensive degree of shaping. Also within the category of flaked pieces, multifacial cores and polyhedrons are intensively flaked cores with more than three flaking surfaces, with polyhedrons generally regarded as approaching spherical forms. However, using spherical forms as the distinguish standard is still subjective. In this study, artifacts with a thickness-to-length ratio greater than 0.6 are operationally classified as polyhedrons (n = 17), as these specimens tend to exhibit more comparable dimensions along the three principal axes, whereas the remaining pieces are classified as multifacial cores (n = 12). Importantly, this classification reflects a morphological continuum rather than discrete categories; the threshold value of 0.6 is adopted as a practical criterion to distinguish polyhedrons from multifacial cores, producing roughly comparable sample sizes for the two categories within the assemblage.

@@

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Cores",
                      "spheroid" = "Sub/ Spheroids",
                      "multi_poly" = "Polyhedrons/ Multifacials"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Cores" = "gray", 
    "Sub/ Spheroids" = "orange", 
    "Polyhedrons/ Multifacials" = "#b2abd2"
  )) +
  theme_minimal(base_size = 10) +
  theme(
    # 图例放在右上角
    legend.position = c(0.95, 0.95),
    legend.justification = c(1, 1),
    legend.background = element_rect(fill = "white", color = "black", size = 0.3),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.1, "cm"),
    legend.margin = margin(3, 3, 3, 3),
    legend.title = element_text(size = 9),  # 调整图例标题大小

    # 其他样式
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9)
  )

# 显示图形
print(p_dimensions)

@@

ggsave(
  filename = here("analysis/paper/figures/basic_characteristics2.png"),
  plot = p_combined,
  width = 9,   
  height = 4,  
  dpi = 300
)

knitr::include_graphics(here("analysis/paper/figures/basic_characteristics2.png"))
```

@@

```{r, echo=FALSE}
#| label: fig-benn-ternary-plot
#| fig-cap: "Ternary plot showing flake scar orientation analysis"
library(dplyr)
library(ggtern)

@@

benn.output <- benn.output %>%
  mutate(Z = 1 - EL - IS)

# 计算凸包函数
get_hull <- function(df) {
  if(nrow(df) < 3) return(df)  
  hull_idx <- chull(df$Z, df$IS)
  df[hull_idx, ]
}

# 为每个类型生成凸包
hull_data <- benn.output %>%
  group_by(type) %>%
  do(get_hull(.)) %>%
  ungroup()

# 定义颜色映射
type_colors <- c(
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99", 
  "subspheroid" = "#fdb863",
  "spheroid" = "#e66101"
)

fill_colors <- c(
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99", 
  "subspheroid" = "#fdb863",
  "spheroid" = "#e66101"
)

# 绘图
p_benn_ternary <- ggtern(data = benn.output, aes(x = Z, y = IS, z = EL)) +
  # 添加三角网格掩膜
  geom_mask() +

  # 凸包填充 - 添加在半透明层
  geom_polygon(
    data = hull_data,
    aes(fill = type, group = type),
    color = NA,  # 无边框
    alpha = 0.2  # 填充透明度
  ) +

  # 凸包边框
  geom_polygon(
    data = hull_data,
    aes(color = type, group = type),
    fill = NA,
    size = 0.8,
    alpha = 0.8
  ) +

  # 散点
  geom_point(
    aes(color = type),
    size = 3,
    alpha = 0.8
  ) +

  # 主题和标签
  theme_bw() +
  theme_showarrows() +
  theme_clockwise() +

  # 轴标签
  labs(
    x = "Planar",
    xarrow = "Planarity",
    y = "Isotropic", 
    yarrow = "Isotropy",
    z = "Linear",
    zarrow = "Elongation",
    color = "Type",
    fill = "Type"
  ) +

  # 颜色映射
  scale_color_manual(
    values = type_colors,
    limits = names(type_colors)
  ) +

  # 填充颜色映射
  scale_fill_manual(
    values = fill_colors,
    limits = names(fill_colors)
  ) +

  theme(
    legend.position = "right",
    legend.box = "vertical",
    legend.key = element_rect(fill = "white")
  )

# 保存图形
output_dir <- here("analysis", "paper", "figures")
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

ggsave(
  filename = file.path(output_dir, "benn_ternary_plot.jpg"),
  plot = p_benn_ternary,
  width = 10,
  height = 6,
  dpi = 300,
  bg = "white"
)

# 在文档中显示图形
knitr::include_graphics(file.path(output_dir, "benn_ternary_plot.jpg"))
